---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#",
  fig.path = "tools/"
)
```

# TITAN2

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/TITAN2)](https://cran.r-project.org/package=TITAN2)
<!-- badges: end -->

__TITAN2__ is the second R implementation of Threshold Indicator Taxa ANalysis. It is an R package source controlled with Git on GitHub and [distributed on CRAN](https://cran.r-project.org/package=TITAN2).

To learn more about __TITAN2__, install the package (see below), load it with `library("TITAN2")`, and read the vignette with `vignette("titan2-intro")`.



## Installation

You can install __TITAN2__ in either of two ways. At the present time, we recommend installing __TITAN2__ from GitHub, as it has several new features, e.g. `plot_taxa_ridges()`.


* From Github (dev version): 
```{r, eval=FALSE}
if (!requireNamespace("devtools")) install.packages("devtools")
devtools::install_github("dkahle/TITAN2")
```


* From CRAN: 
```{r, eval=FALSE}
install.packages("TITAN2")
```




## Acknowledgements

This work continues to be supported by the [Department of Geography and Environmental Systems (UMBC)](https://ges.umbc.edu/), [Department of Biology (Baylor)](https://www.baylor.edu/biology/), and [Department of Statistical Science (Baylor)](http://www.baylor.edu/statistics/). 
